### 2021-05-14 content/80.formal-definition.tex

%% TODO: Convert to LaTex math mode. See GitHub rendering for original Unicode symbols
% https://github.com/stain/ro-crate-paper/blob/master/content/80.formal-definition.md

% Text remains for now as Markdown

@@

    $\mathbb{L}$ro-crate = { Property(p), Class(c), Literal(x), ℝ, $\mathbb{S}$ }
    $\mathbb{D}$ =  $\mathbb{Iri}$
    $\mathbb{Iri}$ ≡  { IRIs as defined in RFC3987 }
    ℝ ≡  { real or integer numbers }
    $\mathbb{S}$ ≡  { literal strings }

\begin{markdown}
The domain of discourse is the set of $\mathbb{Iri}$ identifiers [@doi:10.17487/rfc3987] (notation `<http://example.com/>`), with additional descriptions using numbers ℝ (notation `13.37`) and literal strings $\mathbb{S}$ (notation `“Hello”`).

From this formalised language `$\mathbb{L}$ro-crate` an RO-Crate can be interpreted in any representation that can gather these descriptions, their properties, classes, and literal attributes.  
\end{markdown}

@@

Below is using `$\mathbb{L}$ro-crate` to define a minimal RO-Crate:
\end{markdown}

    RO-Crate(R) ⊨  Root(R) ∧ Mentions(R, R) ∧
        hasPart(R, d) ∧ Mentions(R, d) ∧ DataEntity(d) ∧
        Mentions(R, c) ∧ ContextualEntity(c)
    ∀r Root(r) →  Dataset(r) ∧ name(r, n)∧ description(r, d)
                            ∧ published(r, date) ∧ license(e, l)
    ∀e∀n name(e, n) →  Literal(n)
    ∀e∀d description(e, d) →  Literal(d)
    ∀e∀date datePublished(e, date) →  Literal(date)
    ∀e∀l license(e, l) →  ContextualEntity(l)
    DataEntity(e) ≡  File(e) ⊕ Dataset(e)
    Entity(e) ≡  DataEntity(e) ∨ ContextualEntity(e)
    ∀e Entity(e) → Class(e)
    Mentions(R, s) ⊨  Relation(s, p, e)  ⊕  Attribute(s, p, l)
    Relation(s, p, o) ⊨ Entity(s) ∧ Property(p) ∧ Entity(o)
    Attribute(s, p, x) ⊨ Entity(s)  ∧ Property(p) ∧ Literal(x)
    Literal(x) ≡  x ∈ ℝ  ⊕   x ∈  $\mathbb{S}$

\begin{markdown}
An `RO-Crate(R)` is defined as a self-described _Root Data Entity_, which describes and contains parts (_data entities_), which are further described in _contextual entities_.  These terms align with their use in the [RO-Crate 1.1 terminology](https://www.researchobject.org/ro-crate/1.1/terminology).

The `Root(r)` is a type of `Dataset(r)`, and must have the metadata to literal attributes to provide a `name`, `description` and `datePublished`, as well as a contextual entity identifying its license. These predicates correspond to the RO-Crate 1.1 [requirements for the root data entity](https://www.researchobject.org/ro-crate/1.1/root-data-entity.html#direct-properties-of-the-root-data-entity).

The concept of an `Entity(e)` is introduced as being either a DataEntity(e), a `ContextualEntity(e)`, or [both](https://www.researchobject.org/ro-crate/1.1/contextual-entities.html#contextual-vs-data-entities); and must be typed with at least one `Class(e)`.

For simplicity in this formalization (and to assist production rules below) `R` is a constant representing a single RO-Crate, typically written to independent RO-Crate Metadata files. `R` is used by `Mentions(R, e)` to indicate that `e` is an Entity described by the RO-Crate and therefore its metadata (a set of Relation and Attribute predicates) form part of the RO-Crate serialization. `Relation(s, p, o)` and `Attribute(s, p, x)` are defined as a _subject-predicate-object_ triple pattern from an `Entity(s)` using a `Property(p)` to either another `Entity(o)` or a `Literal(x)` value.
\end{markdown}

@@

The below is an example RO-Crate represented using the above formalization, assuming a base URI of `http://example.com/ro/123/`:
\end{markdown}

    RO-Crate(<http://example.com/ro/123/>)
    name(<http://example.com/ro/123/, 
        “Data files associated with the manuscript:Effects of …”)
    description(<http://example.com/ro/123/, 
        “Palliative care planning for nursing home residents …")
    datePublished(<http://example.com/ro/123/>, “2017")
    license(<http://example.com/ro/123/>, 
        <https://creativecommons.org/licenses/by-nc-sa/3.0/au/>
    ContextualEntity(<https://creativecommons.org/licenses/by-nc-sa/3.0/au/>)
    name(<https://creativecommons.org/licenses/by-nc-sa/3.0/au/, 
        “Attribution-NonCommercial-ShareAlike 3.0 Australia (CC BY-NC-SA 3.0 AU)”)

    hasPart(<http://example.com/ro/123/>, <http://example.com/ro/123/file.txt>)
    File(<http://example.com/ro/123/survey.csv>)
    name(<http://example.com/ro/123/survey.csv>, “Survey of care providers”)
    hasPart(<http://example.com/ro/123/>, <http://www.example.om/ro/123/folder/>)
    Dataset(<http://example.com/ro/123/interviews/>)
    name(<http://example.com/ro/123/interviews/>, 
        “Audio recordings of care provider interviews”)

@@

### Mapping to RDF with schema.org

@@

Note that in the JSON-LD serialization of RO-Crate the expression of `Class` and `Property `is typically indirect, as the JSON-LD `@context` maps to schema.org IRIs, which when resolved as Linked Data embeds their formal definition as RDFa.

@@

While in the conceptual model the _RO-Crate Metadata File_ can be seen as the top-level node that describes the _RO-Crate Root_, in the formal model (and the JSON-LD format) the metadata file descriptor is an additional contextual entity and not affecting the depth-limit of the RO-Crate.

### Forward-chained Production Rules for JSON-LD

@@

This exposes the first order logic domain of discourse of IRIs, with rational numbers and strings as their corresponding JSON-LD representation. These production rules first grow the graph of `R` by adding a transitive rule that anything described in `R` which is related to `o` means that `o` is also mentioned by the RO-Crate `R`. For simplicity this rule is one-way; in practice the JSON-LD graph can also contain free-standing contextual entities that have outgoing relations to data- and contextual entities.

[^2]:
    Limitations: The full list of types, relations and attribute properties from the RO-Crate specification are not included. Examples shown include `datePublished`, `CreativeWork` and `name`. Contextual entities not related from the RO-Crate (e.g. using inverse relations to a data entity) would not be covered by the single direction `Mentions(R, s)` production rule; see [issue #122](https://github.com/ResearchObject/ro-crate/issues/122). The `datePublished(e, date)` rule do not include syntax checks for the ISO 8601 datetime format. Compared with RO-Crate examples, this generated JSON-LD does not use a `@context` as the IRIs are produced unshortened; a post-step could do JSON-LD Flattening with a versioned RO-Crate context.

>>>>

%% TODO: Being onverted to LaTex math mode. See GitHub rendering for original Unicode symbols
% https://github.com/stain/ro-crate-paper/blob/master/content/80.formal-definition.md

% Text remains for now as Markdown blocks

@@

Definition of language $\mathcal{L}_{rocrate}$:

\begin{eqnarray*}
    \mathcal{L}_{rocrate}   & \equiv & { Property(p), Class(c), 
                            Literal(x), \mathbb{R}, \mathbb{S} } \\
    \mathbb{D}              & \equiv & \mathbb{IRI} \\
    \mathbb{IRI}            & \equiv & { \text{IRIs as defined in RFC3987} } \\
    \mathbb{R}              & \equiv & { \text{real or integer numbers} } \\
    \mathbb{S}              & \equiv & { \text{literal strings} }
\end{eqnarray*}

\begin{markdown}
The domain of discourse $\mathbb{D}$ is the set of $\mathbb{IRI}$ identifiers [@doi:10.17487/rfc3987] (notation `<http://example.com/>`), with additional descriptions using numbers $\mathbb{R}$ (notation `13.37`) and literal strings $\mathbb{S}$ (notation `"Hello"`).

From this formalised language $\mathcal{L}_{rocrate}$ we can interpret an RO-Crate in any representation that can gather these descriptions, their properties, classes, and literal attributes.  
\end{markdown}

@@

Below we use $\mathcal{L}_{rocrate}$ to define a minimal RO-Crate:
\end{markdown}

\begin{eqnarray*}
ROCrate(R)                                 & \models & Root(R) \land Mentions(R, R) \land hasPart(R, d) \land \\
                                            & & Mentions(R, d) \land DataEntity(d) \land \\
                                            & & Mentions(R, c) \land ContextualEntity(c) \\
\forall r Root(r)                           & \Rightarrow & Dataset(r) \land name(r, n) \land description(r, d) \land \\
                                            & &             published(r, date) \land license(e, l) \\
\forall e \forall n \ name(e, n)            & \Rightarrow & Literal(n) \\
\forall e \forall s \ description(e, s)     & \Rightarrow & Literal(s) \\
\forall e \forall d \ datePublished(e, d)   & \Rightarrow & Literal(d) \\
\forall e \forall l \ license(e, l)         & \Rightarrow & ContextualEntity(l) \\
DataEntity(e)                               & \equiv &      File(e) \oplus Dataset(e) \\
Entity(e)                                   & \equiv &      DataEntity(e) \lor ContextualEntity(e) \\
\forall e \ Entity(e)                       & \Rightarrow & Class(e) \\
Mentions(R, s)                              & \models &     Relation(s, p, e) \oplus Attribute(s,  p, l) \\
Relation(s, p, o)                           & \models &     Entity(s) \land Property(p) \land  Entity(o) \\
Attribute(s, p, x)                          & \models &     Entity(s) \land Property(p) \land Literal(x) \\
Literal(x)                                  & \equiv &      x \in \mathbb{R} \oplus x \in \mathbb{S}
\end{eqnarray*}

\begin{markdown}
An $ROCrate(R)$ is defined as a self-described _Root Data Entity_, which describes and contains parts (_data entities_), which are further described in _contextual entities_.  These terms align with their use in the [RO-Crate 1.1 terminology](https://www.researchobject.org/ro-crate/1.1/terminology).

The $Root(r)$ is a type of $Dataset(r)$, and must have the metadata to literal attributes to provide a $name$, $description$ and $datePublished$, as well as a contextual entity identifying its license. These predicates correspond to the RO-Crate 1.1 [requirements for the root data entity](https://www.researchobject.org/ro-crate/1.1/root-data-entity.html#direct-properties-of-the-root-data-entity).

The concept of an $Entity(e)$ is introduced as being either a $DataEntity(e)$, a $ContextualEntity(e)$, or [both](https://www.researchobject.org/ro-crate/1.1/contextual-entities.html#contextual-vs-data-entities); and must be typed with at least one $Class(e)$.

For simplicity in this formalization (and to assist production rules below) $R$ is a constant representing a single RO-Crate, typically written to independent RO-Crate Metadata files. $R$ is used by $Mentions(R, e)$ to indicate that $e$ is an Entity described by the RO-Crate and therefore its metadata (a set of $Relation$ and $Attribute$ predicates) form part of the RO-Crate serialization. $Relation(s, p, o)$ and $Attribute(s, p, x)$ are defined as a _subject-predicate-object_ triple pattern from an $Entity(s)$ using a $Property(p)$ to either another $Entity(o)$ or a $Literal(x)$ value.
\end{markdown}

@@

The below is an example RO-Crate represented using the above formalization, assuming a base URI of `<http://example.com/ro/123/>`:

    ROCrate(<http://example.com/ro/123/>)
    name(<http://example.com/ro/123/, 
        “Data files associated with the manuscript:Effects of …”)
    description(<http://example.com/ro/123/, 
        “Palliative care planning for nursing home residents …")
    datePublished(<http://example.com/ro/123/>, “2017")
    license(<http://example.com/ro/123/>, 
        <https://creativecommons.org/licenses/by-nc-sa/3.0/au/>
    ContextualEntity(<https://creativecommons.org/licenses/by-nc-sa/3.0/au/>)
    name(<https://creativecommons.org/licenses/by-nc-sa/3.0/au/, 
        “Attribution-NonCommercial-ShareAlike 3.0 Australia (CC BY-NC-SA 3.0 AU)”)

    hasPart(<http://example.com/ro/123/>, <http://example.com/ro/123/file.txt>)
    File(<http://example.com/ro/123/survey.csv>)
    name(<http://example.com/ro/123/survey.csv>, “Survey of care providers”)
    hasPart(<http://example.com/ro/123/>, <http://www.example.om/ro/123/folder/>)
    Dataset(<http://example.com/ro/123/interviews/>)
    name(<http://example.com/ro/123/interviews/>, 
        “Audio recordings of care provider interviews”)
\end{markdown}

@@

\begin{markdown}
### Mapping to RDF with schema.org

@@

Note that in the JSON-LD serialization of RO-Crate the expression of $Class$ and $Property $is typically indirect, as the JSON-LD $@context$ maps to schema.org IRIs, which when resolved as Linked Data embeds their formal definition as RDFa. 
\end{markdown}

\begin{markdown}

@@

While in the conceptual model the _RO-Crate Metadata File_ can be seen as the top-level node that describes the _RO-Crate Root_, in the formal model (and the JSON-LD format) the metadata file descriptor is an additional contextual entity and not affecting the depth-limit of the RO-Crate.
\end{markdown}

\begin{markdown}
### Forward-chained Production Rules for JSON-LD

@@

This exposes the first order logic domain of discourse of IRIs, with rational numbers and strings as their corresponding JSON-LD representation. These production rules first grow the graph of $R$ by adding a transitive rule – anything described in $R$ which is related to $o$, means that $o$ is also mentioned by the $ROCrate(R)$. For simplicity this rule is one-way; in practice the JSON-LD graph can also contain free-standing contextual entities that have outgoing relations to data- and contextual entities.

[^2]:
    Limitations: The full list of types, relations and attribute properties from the RO-Crate specification are not included. Examples shown include $datePublished$, $CreativeWork$ and $name$. Contextual entities not related from the RO-Crate (e.g. using inverse relations to a data entity) would not be covered by the single direction $Mentions(R, s)$ production rule; see [issue 122](https://github.com/ResearchObject/ro-crate/issues/122). The $datePublished(e, date)$ rule do not include syntax checks for the ISO 8601 datetime format. Compared with RO-Crate examples, this generated JSON-LD does not use a $@context$ as the IRIs are produced unshortened; a post-step could do JSON-LD Flattening with a versioned RO-Crate context.
\end{markdown}
